Tsbrowse: a genome browser for Ancestral Recombination Graphs

### 2024-12-12 manuscript/paper.tex
fixup citations a bit

Biobank-driven research has accelerated the development of several
technologies, including the ability to infer Ancestral Recombination Graphs
(ARGs) from large-scale genetic data
\citep{Kelleher2019,Speidel2019,Zhang2023,Gunnarsson2024.08.31.610248,Deng2024.03.16.585351}.
ARGs encode the interwoven paths of genetic inheritance resulting from
recombination, and can be represented as genetic genealogies along the sequence
\citep{Wong2024,Hudson1983,Griffiths1997}. They capture rich detail about
ancestral processes influencing the samples, and have garnered great interest
in several applications in population and statistical genetics. These include
the detection of ultra-rare genetic variation \citep{Zhang2023}, the mapping of
quantitative trait loci \citep{Link2023}, and the sparse and efficient modeling
of linkage disequilibrium \citep{Nowbandegani2023}. It is evident that ARG
inference is increasingly becoming a part of genetic data analysis pipelines.
Yet, ARG visualisation, which is crucial for downstream analyses remains
challenging due to the complexity of inferences from empirical data. Existing
visualisation tools (see \hyperref[sec:Discussion]{Discussion}) focus on the
semantic representation of ARGs, but such representations have limited utility
in informing downstream analyses.

To address this gap, we have developed \texttt{tsbrowse}, an open-source app
intended as a genome browser for ARGs. Its  focus is on visualising the
information structure in ARGs, thereby improving  inference and analysis at
biobank-scale. \texttt{Tsbrowse} provides interactive visual summaries of  the
fundamental properties of ARGs, i.e. the mutations, nodes, edges and trees they
encode. \texttt{Tsbrowse} operates on ARGs encoded as succinct tree sequences
\citep{Wong2024}. This is the native format of \texttt{tskit} \citep{tskit}, a
widely-used library for interchanging ARGs inferred using a variety of methods
\citep{Speidel2019,Zhang2023,Haller2019,Baumdicker2021,Gunnarsson2024.08.31.610248,Deng2024.03.16.585351}.
The interface is run on a web browser. It works on biobank-scale datasets
without requiring \textcolor{red}{raw data uploads to a server}. This is ideal
for performing quality control and exploratory analyses on ARGs inferred from
biobank datasets. \section{Results} \subsection{Architecture} \texttt{Tsbrowse}
is run in a step-wise approach (Figure \ref{fig:Figure_1}), which makes it
convenient to handle large datasets. The initial step (\texttt{tsbrowse
preprocess}) prepares the data for visualisation. It can be performed
separately on a large machine, and the output stored for later use. The second
step (\texttt{tsbrowse serve}), which can be run on a smaller machine, creates
and renders visualisations. This step follows a client-server architecture. On
the server side (the machine on which \texttt{tsbrowse serve} is run), the
Holoviz suite of libraries \citep{Holoviz} is leveraged to create
visualisations based on the pre-processed data. This choice of architecture is
prompted by the need to represent properties of large-scale ARGs in a
meaningful and performant way. For example, a visual representation of millions
of mutations in SARS-Cov-2 ARGs \citep{Zhan2023.06.08.544212} could quickly
lead to performance bottlenecks and overplotting. Datashader
\citep{Datashader}, a component Holoviz package, overcomes this challenge by
rasterising huge datasets (see \hyperref[subsec:User_Interface]{User Interface}
for details, and Supplementary Figure \ref{fig:Supplementary_Figure_4} for a
tsbrowse view of SARS-Cov-2 mutations). A web browser client connects to the
server to render and interact with these visualisations. Heavy data processing
is limited to the server side, keeping the client-side interface lightweight.
For large datasets with complex visualisations, this achieves a smoother user
experience.

@@

\subsection{Data model} \label{subsec:Data_Model} The foundation of
\texttt{tsbrowse} is the \texttt{tskit} data model. As \texttt{tsbrowse} is
based on this generalised ARG representation \citep{Wong2024} it can be applied
to ARGs from several widely-used inference methods including \texttt{Relate}
\citep{Speidel2019}, \texttt{ARG-Needle} \citep{Zhang2023} and \texttt{SINGER}
\citep{Deng2024.03.16.585351}. In \texttt{tskit}, ARGs are encoded as a
collection of tables, each of which store information about a specific ARG
property. For example, the Node table describes sampled or ancestral sequences
and records their time of birth, and the individual and population they belong
to. The Edge table describes the relationship between pairs of nodes, and
records the node IDs and the genomic locations over which the edge spans. The
Mutation table records mutation events in the history of the sample, including
the change of allelic state, time of event and node in which the event
occurred. The Node and Edge tables describe the genealogical relationships
between the sampled and ancestral sequences, while the Mutation table describes
sequence variation contained in the samples \citep{Ralph2020}. This compact
representation enables the efficient storage and querying of biobank-scale
ARGs. \texttt{Tsbrowse} augments these tables with additional columns to create
a compressed \texttt{.tsbrowse} file, containing all necessary semantic
metadata required for automatic visualisation (see
\hyperref[subsec:User_Interface]{User Interface}).

@@

In current practice, simulated ARGs are used as ground-truth to validate
inferences from empirical data
\citep{Kelleher2019,Zhang2023,Speidel2019,Gunnarsson2024.08.31.610248}.
Sophisticated techniques
\citep{Haller2019,Baumdicker2021,Adrion2020,Lauterbur2023} to simulate ARGs
under complex, realistic demographic models \cite{Anderson-Trocme2023} now
exist. However, models of simulation do not capture \textit{all} the nuances of
population datasets. This results in a clear gap between ARG inference and
application. \texttt{Tsbrowse} captures ARG properties in a model-free way,
providing the means to qualitatively assess inferences. With notable features
of browsability along the genome, scalability and applicability to diverse data
sources, inference and simulation methods, \texttt{tsbrowse} has the potential
to be integrated into biobank-scale data pipelines to improve ARG inference and
analysis.

@@

\subsection{Inference of selective sweep dataset (Suppl. Fig.
\ref{fig:Supplementary_Figure_1})} The following software were used to infer
ARGs from the truth dataset described above: \texttt{tsinfer version 0.3.3}
\citep{Kelleher2019}, \texttt{tsdate version 0.2.1}, \texttt{Relate version
1.2.2} \cite{Speidel2019}, \texttt{ARG-needle version 1.0.3} \citep{Zhang2023},
\texttt{SINGER version 0.1.8-beta} \citep{Deng2024.03.16.585351}. For all
methods, the following parameters of inference were used: \textit{recombination
rate = 1e-8}, \textit{mutation rate = 1e-8}, \textit{effective population size
= 10,000}. Default values were used for all other inference parameters.

>>>>

Ancestral recombination graphs (ARGs) describe how a set of sampled sequences 
are related to each other at each position of the genome in a recombining
species~\citep{brandt2024promise,lewanski2024era,nielsen2024inference,
wong2024general}, and there has been significant recent progress 
on inference through a range of approaches~\citep{rasmussen2014genome,
speidel2019method,kelleher2019inferring,zhang2023biobank,gunnarsson2024scalable,
deng2024robust}.

Biobank-driven research has accelerated the development of several
technologies, including the ability to infer Ancestral Recombination Graphs
(ARGs) from large-scale genetic data
% \citep{Kelleher2019,Speidel2019,Zhang2023,Gunnarsson2024.08.31.610248,Deng2024.03.16.585351}.
ARGs encode the interwoven paths of genetic inheritance resulting from
recombination, and can be represented as genetic genealogies along the sequence
\citep{wong2024general,Hudson1983,Griffiths1997}. They capture rich detail about
ancestral processes influencing the samples, and have garnered great interest
in several applications in population and statistical genetics. These include
the detection of ultra-rare genetic variation \citep{zhang2023biobank}, the mapping of
quantitative trait loci \citep{Link2023}, and the sparse and efficient modeling
of linkage disequilibrium \citep{Nowbandegani2023}. It is evident that ARG
inference is increasingly becoming a part of genetic data analysis pipelines.
Yet, ARG visualisation, which is crucial for downstream analyses remains
challenging due to the complexity of inferences from empirical data. Existing
visualisation tools (see \hyperref[sec:Discussion]{Discussion}) focus on the
semantic representation of ARGs, but such representations have limited utility
in informing downstream analyses.

To address this gap, we have developed \texttt{tsbrowse}, an open-source app
intended as a genome browser for ARGs. Its  focus is on visualising the
information structure in ARGs, thereby improving  inference and analysis at
biobank-scale. \texttt{Tsbrowse} provides interactive visual summaries of  the
fundamental properties of ARGs, i.e. the mutations, nodes, edges and trees they
encode. \texttt{Tsbrowse} operates on ARGs encoded as succinct tree sequences
\citep{wong2024general}. This is the native format of \texttt{tskit} \citep{tskit}, a
widely-used library for interchanging ARGs inferred using a variety of methods
\citep{speidel2019method,zhang2023biobank,Haller2019,Baumdicker2021,gunnarsson2024scalable,
deng2024robust}.
The interface is run on a web browser. It works on biobank-scale datasets
without requiring \textcolor{red}{raw data uploads to a server}. This is ideal
for performing quality control and exploratory analyses on ARGs inferred from
biobank datasets. \section{Results} \subsection{Architecture} \texttt{Tsbrowse}
is run in a step-wise approach (Figure \ref{fig:Figure_1}), which makes it
convenient to handle large datasets. The initial step (\texttt{tsbrowse
preprocess}) prepares the data for visualisation. It can be performed
separately on a large machine, and the output stored for later use. The second
step (\texttt{tsbrowse serve}), which can be run on a smaller machine, creates
and renders visualisations. This step follows a client-server architecture. On
the server side (the machine on which \texttt{tsbrowse serve} is run), the
Holoviz suite of libraries \citep{Holoviz} is leveraged to create
visualisations based on the pre-processed data. This choice of architecture is
prompted by the need to represent properties of large-scale ARGs in a
meaningful and performant way. For example, a visual representation of millions
of mutations in SARS-Cov-2 ARGs \citep{Zhan2023.06.08.544212} could quickly
lead to performance bottlenecks and overplotting. Datashader
\citep{Datashader}, a component Holoviz package, overcomes this challenge by
rasterising huge datasets (see \hyperref[subsec:User_Interface]{User Interface}
for details, and Supplementary Figure \ref{fig:Supplementary_Figure_4} for a
tsbrowse view of SARS-Cov-2 mutations). A web browser client connects to the
server to render and interact with these visualisations. Heavy data processing
is limited to the server side, keeping the client-side interface lightweight.
For large datasets with complex visualisations, this achieves a smoother user
experience.

@@

\subsection{Data model} \label{subsec:Data_Model} The foundation of
\texttt{tsbrowse} is the \texttt{tskit} data model. As \texttt{tsbrowse} is
based on this generalised ARG representation \citep{wong2024general} it can be applied
to ARGs from several widely-used inference methods including \texttt{Relate}
\citep{speidel2019method}, \texttt{ARG-Needle} \citep{zhang2023biobank} 
and \texttt{SINGER} \citep{deng2024robust}. In \texttt{tskit}, ARGs are encoded as a
collection of tables, each of which store information about a specific ARG
property. For example, the Node table describes sampled or ancestral sequences
and records their time of birth, and the individual and population they belong
to. The Edge table describes the relationship between pairs of nodes, and
records the node IDs and the genomic locations over which the edge spans. The
Mutation table records mutation events in the history of the sample, including
the change of allelic state, time of event and node in which the event
occurred. The Node and Edge tables describe the genealogical relationships
between the sampled and ancestral sequences, while the Mutation table describes
sequence variation contained in the samples \citep{Ralph2020}. This compact
representation enables the efficient storage and querying of biobank-scale
ARGs. \texttt{Tsbrowse} augments these tables with additional columns to create
a compressed \texttt{.tsbrowse} file, containing all necessary semantic
metadata required for automatic visualisation (see
\hyperref[subsec:User_Interface]{User Interface}).

@@

In current practice, simulated ARGs are used as ground-truth to validate
inferences from empirical data
\citep{kelleher2019inferring,zhang2023biobank,speidel2019method,
gunnarsson2024scalable}.
Sophisticated techniques
\citep{Haller2019,Baumdicker2021,Adrion2020,Lauterbur2023} to simulate ARGs
under complex, realistic demographic models \cite{Anderson-Trocme2023} now
exist. However, models of simulation do not capture \textit{all} the nuances of
population datasets. This results in a clear gap between ARG inference and
application. \texttt{Tsbrowse} captures ARG properties in a model-free way,
providing the means to qualitatively assess inferences. With notable features
of browsability along the genome, scalability and applicability to diverse data
sources, inference and simulation methods, \texttt{tsbrowse} has the potential
to be integrated into biobank-scale data pipelines to improve ARG inference and
analysis.

@@

\subsection{Inference of selective sweep dataset (Suppl. Fig.
\ref{fig:Supplementary_Figure_1})} The following software were used to infer
ARGs from the truth dataset described above: \texttt{tsinfer version 0.3.3}
\citep{kelleher2019inferring}, \texttt{tsdate version 0.2.1}, 
\texttt{Relate version 1.2.2} \cite{speidel2019method}, 
\texttt{ARG-needle version 1.0.3} \citep{zhang2023biobank},
\texttt{SINGER version 0.1.8-beta} \citep{deng2024robust}. 
For all
methods, the following parameters of inference were used: \textit{recombination
rate = 1e-8}, \textit{mutation rate = 1e-8}, \textit{effective population size
= 10,000}. Default values were used for all other inference parameters.
